---
title: "Assignment-1"
author: "Cory Saunders"
date: "August 15, 2016"
output: html_document
#output: md_document
---

### Coursera Data Science Specialization
### Reproducible Research: assignment 1

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE, results='hide'}
library(knitr)
library(ggplot2)
library(dplyr)
library(Hmisc)
```

#### Loading and preprocessing the data
##### 1. Load the data

If the data has not been dowloaded to this projects location, it can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip. Simply download the link and unzip the file.
```{r, results='markup', warning=TRUE, message=TRUE}
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}
activityData <- read.csv('activity.csv')
```

##### Code to prepare dataset and/or processing the data

```{r}
stepsByDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
```


##### 2. Make a histogram of the total number of steps taken each day

```{r}
hist(stepsByDay, 10, main="Total steps by day", xlab="")
```

##### 3. Mean and median number of steps taken each day

```{r}
stepsByDayMean <- mean(stepsByDay)
stepsByDayMedian <- median(stepsByDay)
```

* Mean steps per day: `r stepsByDayMean`
* Median steps per day: `r stepsByDayMedian`

##### 4. Time series plot of the average number of steps taken

```{r}
dailyAverageSteps <- tapply(activityData$steps, activityData$interval, mean, na.rm = TRUE, simplify = T)
df.dailyAverageSteps <- data.frame(interval=as.integer(names(dailyAverageSteps)), avgStep=dailyAverageSteps)

averageStepsPerTimeBlock <-
  aggregate(x=list(meanSteps=activityData$steps), 
  by=list(interval=activityData$interval), FUN=mean,
  na.rm=TRUE)

ggplot(df.dailyAverageSteps, aes(x=interval, y=avgStep)) +
    geom_line()+
    labs(title = "Time series plot of average steps in different intervals", xlab("5-minute interval"), ylab("Average steps"))
```

##### 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r}
mostSteps <- which.max(averageStepsPerTimeBlock$meanSteps)
timeMostSteps <- gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", averageStepsPerTimeBlock[mostSteps,'interval'])
```

* Most steps at: `r timeMostSteps`

##### 6. Code to describe and show a strategy for imputing missing data
  
```{r}
numMissingValues <- length(which(is.na(activityData$steps)))
```

* Number of missing values: `r numMissingValues`

##### Fill in missing values with the mean number of steps taken in the same 5 minute period

```{r}
activityDataImputed <- activityData
activityDataImputed$steps <- impute(activityDataImputed$steps, fun=mean)
```

##### 7. Make a histogram of the total number of steps taken each day, for the modified data

```{r}
stepsByDayImputed <- tapply(activityDataImputed$steps, activityDataImputed$date, sum)
df.stepsByDayImputed <- as.data.frame(stepsByDayImputed)

qplot(stepsByDayImputed, xlab='Total steps per day (Imputed)', ylab='Frequency')
```

##### Calculate and report the mean and median total number of steps taken per day after gaps have been filled with daily mean

```{r}
stepsByDayImputedMean <- mean(stepsByDayImputed)
stepsByDayImputedMedian <- median(stepsByDayImputed)
```

* Mean (Imputed): `r stepsByDayImputedMean`
* Median (Imputed):  `r stepsByDayImputedMedian`

##### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

First, in the dataset, separate weekdays and weekends
```{r}
activityDataImputed$dateType <-  ifelse(as.POSIXlt(activityDataImputed$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

Make a graph to compare and see if there is a difference
```{r}
averagedActivityDataImputed <- aggregate(steps ~ interval + dateType, data=activityDataImputed, mean)
ggplot(averagedActivityDataImputed, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```

